Follow-up Samples

### 2017-01-15 submission/manuscript_outline_20161024.tex
add updated pdf

\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
health problem with early detection being used as a key component in
mitigating deaths due to the disease. Previous research suggests a link
between stool bacterial microbiome and CRC. The overall objective was to
investigate the changes in the bacterial microbiome after surgery in
patients with lesion (i.e.~adenoma or carcinoma). Specifically, we
wanted to identify what within the community was different within those
undergoing surgical removal of lesion. We also investigated the use of
the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build
models to either classify individuals as having a lesion or whether,
based on the bacterial microbiome, the sample could be classified
correctly as before or after surgery.\\
\textbf{Results:} Adenoma individual's bacterial microbiome were more
similar to their pre-surgery sample then those with carcinoma (P-value =
0.00198) and this was also reflected in FIT as well (P-value =
2.15e-05). There was no significant difference in any indivdiual OTU
between samples before and after surgery (P-value \textgreater{} 0.125.
A model with a total of 37 variables was able to classify lesion with an
AUC range of 0.847 to 0.791 while the model to classify samples as
before and after had 33 with an AUC range of 0.79 to 0.651 for 100 20
repeated 10-fold cross-validated runs. Both models had a significant
decrease in the positive probability of a lesion between individual's
before versus after surgery samples (P-value = 1.91e-11 and 6.72e-12).
In total there were 14 OTUs that were common to both models and were
mostly commensals with largest representation from OTUs belonging to
Bacteroides, Blautia, Streptococcus, and Clostridiales.\\
\textbf{Conclusions:} Our data suggests that treatment not only
significantly reduces the probability of having a colonic lesion but
also causes detectable changes in the bacterial microbiome. Further
surveillance of these individuals will enable us to determine whether
models such as the one we present here can also be used to predict
recurrence of colorectal cancer.

@@

\textbf{\emph{Bacterial Community and Fit Changes before and after
Treatment}} Based on thetayc distance metrics, comparing the initial to
the follow up samples, there was a significant difference between the
adenoma and carcinoma groups (P-value = 0.00198) {[}Figure 1a{]}. There
was also a significant difference in FIT between initial and follow up
samples with the carcinoma group having a significant decrease in FIT
versus the adenoma group (P-value = 2.15e-05) {[}Figure 1b{]}. The whole
community structure before and after surgery are visualized on NMDS
graphs for both adenoma {[}Figure1c{]} (PERMANOVA = 0.002) and carcinoma
{[}Figure 1d{]} (PERMANOVA = 0.997). When all initial and follow up
samples were compared to each other there was no significant overall
difference between them (PERMANOVA = 0.085). There was also no
significant difference between initial and follow up samples for
observed OTUs, Shannon diversity, or evenness after correction for
multiple comparisons {[}Table S1{]}. There was no significant difference
between initial and follow up samples for any single OTU {[}Figure
S1{]}.

@@

\textbf{\emph{Full and Reduced Model Results}} Since differences were
observed between initial and follow up samples and only a small number
of individuals were positive for previously associated CRC bacteria we
next investigated if we could create a model that could adequately
classify and adjust lesion probability based on the bacterial community
and FIT. The lesion model had an AUC range of 0.723 to 0.795 versus the
initial follow up model which had and AUC range of 0.451 to 0.67 after
100 20 repeated 10-fold cross validations. Interestingly, identification
of the most important variables and reducing the variables considered to
only these increased the AUC in the lesion model (0.791 - 0.847) and
initial and follow up model (0.651 - 0.79).

@@

\textbf{Treatment and Time Differences}

There was no difference in the amount of change in positive probability
for either the full or reduced lesion model for either chemotherapy
(P-value = 0.821 and 0.821) or radiation therapy (P-value = 0.69 and
0.981). Although the initial follow up model was similar there was a
significant decrease in positive probability for those treated with
chemotherapy (P-value = 7.04e-04 and 5.07e-03Time of follow up sample
from initial sampling, did not have a significant difference between
adenoma and carcinoma (uncorrected P-value = 0.784).

@@

In our training set we show that the overall community structure as
measured by different alpha diversity metrics, shows very little change
between controls and those with either adenoma or carcinoma {[}Table
S1{]}. With respect to our test set there was very little difference in
magnitude of change in the thetayc distance metric between those with
adenoma or carcinoma {[}Figure 1a{]}. In contrast, FIT had a large
change in the initial and follow up samples in the carcinoma group
versus the adenoma {[}Figure 1b{]}. An NMDS showed that there was very
little observable change between initial and follow up for the adenoma
group but there was one for the carcinoma group {[}Figure 1c \& 1d{]}.
This cursory information is suggestive that treatment of carcinoma, had
the largest response.

We next created a model that incorporated both patient metadata, FIT,
and the bacterial microbiome to be able to predict lesions (adenoma or
carcinoma). Our middle training model, based on AUC, from 100 80/20
(train/test) splits was similar to the full training data model. It's
10-fold cross validated AUC was similar to it's test set AUC which was
not the case for both the best and worse training model {[}Figure 3{]}.
Using the full training data model we predicted the probability of a
lesion in the inital and follow up samples {[}Figure 4{]}. There was a
signficant decrease in positive probability regardless of whether the
sample was a carcinoma or adenoma. The overall sensitivity for lesion
detection in the intial samples was and for follow ups was . Although
there was a decrease in overall probability of an adenoma or carcinoma
only were below the 0.5 threshold out of the total individuals who were
diagnosed as not having a carcinoma on follow up.

We then investigated which OTUs could potentially be more important in
our model {[}Figure 5 \& Table S3{]}. Many of the OTUs identified
classified to normal flora bacterium {[}Table S3{]}. Only a single OTU
though was significant after multiple comparison correction and the
lowest taxonomic identification of was to . Although there was a
difference in the relative abundance at initial and follow up these
values were not drastically different from the relative abundance values
observed in the control individuals of the training set {[}Figure 5{]}.
Although we were interested in what we could use to classify those with
either adenoma or carcinoma versus normal. We found that the traditional
bacteria associated with CRC were higher in magnitude in the carcinoma
group and there were significant differences in some of these OTUs
between the initial and follow up samples {[}Figure 5 and Tabl S5{]}.
This research provides evidence that it is possible to use bacterial
microbiome data to create a highily sensitive model, that is reactive to
therapy, for detection of adenoma or carcinoma. It accomplishes this by
using a unique sample set in which before and after surgery stool
samples are available for assessment. By using these types of samples we
are not only able to show sensitivity of lesion prediction but also able
to show that this model is reactive. That is to say that after surgery
for removal of the adenoma or carcinoma it decreases the positive
probability to reflect a lower likelihood of the individual having an
adenoma or carcinoma.

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool
(\textbf{insert citation}). Based on previous work by Jobin, et al.
(\textbf{insert citation}) it may not be surprising to see E.coli in the
top 5 OTUs for this model. Similarily, Porphyromonas has also been
implicated in colorectal cancer (\textbf{insert citation}).
Interestingly, many of the other OTUs had taxonomic identification for
resident gut microbes. This could suggest that changes to the resident
microbiome are important to the initiation of adenoma or carcinoma
formation (\textbf{insert citation}) and provide support for the
hypothesis that an initial change in the bacterial microbiome could pave
the way for more inflammatory species: whether by creation of a new
niche for oral microbes (\textbf{insert citation}) or allowing for a
bloom of existing pro-inflammatory residents (\textbf{insert citation}).

Naturally, it is curious that normal staples of many screening studies
such as Fusobacterium, Parvimonas, and Peptostreptococcus were not
present in the majority of the training models. One potential
explanation for this is that FIT provides the same information to the
model as these three organisms and so the model uses FIT preferentially
over them. This has been suggested to be the case in a previous study
(\textbf{insert Baxter Study}). It is also possible that these specific
bacteria play a major role in the progression to carcinoma but may not
be as important in the initiation of an adenoma, which would be
supported by our data {[}Figure 5{]}. Regardless, our study does not
argue against the importance of these bacterium in CRC initiation or
pathogenesis but rather that the model does not utilize these specific
bacteria for prediction purposes. Another potential reason why we did
not identify the ``usual suspects''" is that these bacteria may not
change much between initial and follow up samples in those with an
identified lesion. That is to say that the bacteria are consistently
present even after removal of the lesion by surgery. Finally, it is
likely that within our test set there was not enough indviduals in which
detection was made or relative abundance high enough for these bacteria
to be significant using a paired wilcoxson test.

One limitation in this study is that we do not know whether individuals
in our test set eventually had a subsequent CRC diagnosis. This
information would help to strengthen the case for our Random Forest
based model keeping a number of individuals above the cutoff threshold
even though at follow up they were diagnosed as no longer having a
lesion. Another limitation is that we do not know if adding modern tests
such as the stool DNA test (\textbf{insert citation}) could help improve
our overall AUC. Another limitation is that this study drew heavily from
those with caucasian ancestry. The results may not be immediately
representative of those with either Asian or African ancestry. Finally,
although our training and test set are relatively large we still run the
risk of overfitting or having a model that may not be immediately
extrapolateable to other populations. We've done our best to safeguard
against this by not only running 10-fold cross validation but also
having over 100 different 80/20 splits to try and mimic the type of
variation that might be expected to occur.

By adding patient data such as age, BMI, etc. to the model and showing
that it can successfully help to predict both carcinoma and adenoma our
study provides further data that these patient factors in conjunction
with the bacterial microbiome could potentially influence CRC and
perhaps have a role in formation of adenomas. Further studies need to be
carried out to verify our findings since not only are we dealing with
stool, which could be very different than the communities present on the
actual tissue, but also are dealing with correlations that may not be
representative of the true pathogensis of disease.

Despite these limitations we think that these findings significantly add
to the existing scientific knowledge on CRC and the bacterial
microbiome. The ability for machine learning algorithms to take
bacterial microbiome data and successfully lower positive probability
after either adenoma or carcinoma removal provides evidence that there
are specific signatures associated with these lesions. It also shows
that these algorithms can not only successfully react to successful
treatment regimens but also may be able to one day diagnose CRC with a
high level of accuracy.

@@

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al
{[}13{]}. In brief, study exclusion involved those who had already
undergone surgery, radiation, or chemotherapy, had colorectal cancer
before a baseline stool sample could be obtained, had IBD, a known
hereditary non-polyposis colorectal cancer, or Familial adenomatous
polyposis. Samples used to build the model used for prediction were
collected either prior to a colonoscopy or between 1 - 2 weeks after.
The bacterial microbiome has been shown to nomralize within this time
period {[}18{]}. Kept apart from this training set were a total of 67
individuals that not only had a sample as described previoulsy but also
a follow up sample between 188 - 546 days after surgery and treatment
had been completed. This study was approved by the University of
Michigan Institutional Review Board. All study participants provided
informed consent and the study itself conformed to the guidelines set
out by the Helsinki Declaration.

\textbf{\emph{Fecal Immunochemical Test and 16S rRNA Gene Sequencing}}
FIT was analyzed as previously published using both OC FIT-CHEK and
OC-Auto Micro 80 automated system (Polymedco Inc.) {[}19{]}. 16S rRNA
gene sequencing was completed as previously described by Kozich, et al.
{[}20{]}. In brief, DNA extraction used the 96 well Soil DNA isolation
kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting
system (Eppendorf). The V4 variable region was amplified and the
resulting product was split between three sequencing runs with control,
adenoma, and carcinoma evenly represented on each run. Each group was
randomly assigned to avoid biases based on sample collection location.

\textbf{\emph{Sequence Processing}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described {[}20{]}. The general processing workflow
using mothur is as follows: Paired-end reads were first merged into
contigs, quality filtered, aligned to the SILVA database, screening for
chimeras, classified with a naive Bayesian classifier using the
Ribosomal Database Project (RDP), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm. The numer of sequences for each sample
was rarified to 10521 in an attempt to minimize uneven sampling.

\textbf{\emph{Lesion Model Creation}} The Random Forest {[}21{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) for the 67 individuals with follow up samples.
The model included data on FIT and the bacterial microbiome. Non-binary
data was checked for near zero variance and auto correlation. Data
columns that had near zero variance were removed. Columns that were
correlated with each other over a Spearman correlation coefficient of
0.75 had one of the two columns removed. This pre-processing was
performed with the R package caret (v6.0.73). Optimization of the mtry
hyperparameter involved taking the samples and making 100 80/20
(train/test) splits in the data where control and lesion were equally
represented in the 80 and 20 split, respectively. This 80\% portion was
then split again into an 80/20 split, and run through 20 repeated
10-fold cross validations to optimzie the model's AUC (Area Under the
Curve of the Receiver Operator Characteristic). This resulting model was
then tested on the 20\% of the data that was originally held out from
this overall process. Once the ideal mtry was found the entire 490
sample set was used to create the final Random Forest model on which
testing on the 67-person cohort was completed. The default cutoff of 0.5
was used as the threshold to classify individuals as positive or
negative for lesion. The hyperparameter, mtry, defines the number of
variables to investigate at each split before a new division of the data
is created.

\textbf{\emph{Initial Follow Up Model Creation}} We also investigated
whether a model could be created that could identify before and after
surgery samples. The training set utilized the 67-person cohort that was
previously used for testing of the lesion modeld. The creation of this
model and optimization of the mtry hyperparameter was completed using
the same procedure that was used to create the lesion model.

@@

\textbf{\emph{Statistical Analysis}} The R software package (v3.3.2) was
used for all statisitical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}22{]} in the vegan package
(v2.4.1) while comparisons between ROC curves utilized the method by
DeLong et al. {[}23{]} executed by the pROC (v1.8) package. Comparisons
between probabilities as well as overall amount of OTU between initial
and follow up samples utilized a paired wilcoxson ranked sum test. Where
multiple comparison testing was needed a Benjamini-Hochberg (BH)
correction was applied {[}24{]} and a corrected P-value of less than
0.05 was considered significant. Unless otherwise stated the P-values
reported are those of the BH corrected ones.

\textbf{\emph{Analysis Overview}} Differences in FIT between initial and
follow ups for either adenoma or carcinoma were investigated. Next,
initial and follow up samples were analyzed for differences in alpha and
beta diversity. All OTUs used in the lesion model were also analyzed
using a paired wilcoxson test. The lesion model was then tested for
accuracy in prediction and whether it reduced the positive probability
of lesion after surgery. The most important OTUs for this were used to
build an updated model and this reduced feature model was assessed for
it's similarity to the original model. We then used the initial follow
up model to assess whether this model could classify samples better then
the lesion model. The most important OTUs were then identified from this
model and used to create a reduced feature initial follow up model. This
reduced feature model, as was done with the lesion model, was compared
to the full model for loss of accuracy. Finally, in order to investigate
the relative abundance of specific bacteria, that have been previously
associated with CRC, we selected OTUs that taxonomically classified to
Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus
Assacharolytica, and Porphyromonas Stomatis. Specifically, we wanted to
test if there were any differences based on whether the individual had
an adenoma or carcinoma.

\textbf{\emph{Reproducible methods.}} A detailed and reproducible
description of how the data were processed and analyzed can be found at
\url{https://github.com/SchlossLab/Sze_followUps_2017}.

@@

\hypertarget{ref-obrien_impact_2013}{}
18. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.

\hypertarget{ref-daly_evaluation_2013}{}
19. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
for colorectal cancer screening. Journal of Primary Care \& Community
Health. 2013;4:245--50.

\hypertarget{ref-kozich_development_2013}{}
20. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq Illumina sequencing
platform. Applied and Environmental Microbiology. 2013;79:5112--20.

\hypertarget{ref-breiman_random_2001}{}
21. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
{[}cited 2013 Feb 7{]};45:5--32. Available from:
\href{http://link.springer.com/article/10.1023/A\%3A1010933404324\%20http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}{http://link.springer.com/article/10.1023/A\%3A1010933404324 http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}

\hypertarget{ref-anderson_permanova_2013}{}
22. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
the face of heterogeneous dispersions: What null hypothesis are you
testing? Ecological Monographs {[}Internet{]}. 2013 {[}cited 2017 Jan
5{]};83:557--74. Available from:
\url{http://doi.wiley.com/10.1890/12-2010.1}

\hypertarget{ref-delong_comparing_1988}{}
23. DeLong ER, DeLong DM, Clarke-Pearson DL. Comparing the areas under
two or more correlated receiver operating characteristic curves: A
nonparametric approach. Biometrics. 1988;44:837--45.

\hypertarget{ref-benjamini_controlling_1995}{}
24. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
practical and powerful approach to multiple testing. Journal of the
Royal Statistical Society. Series B (Methodological). 1995;57:289--300.

>>>>

\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
health problem with early detection being used as a key component in
mitigating deaths due to the disease. Previous research suggests a link
between stool bacterial microbiome and CRC. The overall objective was to
investigate the changes in the bacterial microbiome after surgery in
patients with lesion (i.e.~adenoma or carcinoma). Specifically, we
wanted to identify what within the community was different within those
undergoing surgical removal of lesion. We also investigated the use of
the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build
models to either classify individuals as having a lesion or whether,
based on the bacterial microbiome, the sample could be classified
correctly as before or after surgery.\\
\textbf{Results:} Adenoma individual's bacterial microbiome were more
similar to their pre-surgery sample then those with carcinoma (P-value =
0.00198) and this was also reflected in FIT as well (P-value =
2.15e-05). There was no significant difference in any indivdiual OTU
between samples before and after surgery (P-value \textgreater{} 0.125).
A model with a total of 37 variables was able to classify lesion with an
AUC range of 0.847 to 0.791 while the model to classify samples as
before and after had 33 with an AUC range of 0.79 to 0.651 for 100 20
repeated 10-fold cross-validated runs. Both models had a significant
decrease in the positive probability of a lesion between individual's
before versus after surgery samples (P-value = 1.91e-11 and 6.72e-12).
In total there were 14 OTUs that were common to both models and were
mostly commensals with largest representation from OTUs belonging to
Bacteroides, Blautia, Streptococcus, and Clostridiales.\\
\textbf{Conclusions:} Our data suggests that treatment not only
significantly reduces the probability of having a colonic lesion but
also causes detectable changes in the bacterial microbiome. Further
surveillance of these individuals will enable us to determine whether
models such as the one we present here can also be used to predict
recurrence of colorectal cancer.

@@

\textbf{\emph{Bacterial Community and Fit Changes before and after
Treatment}} Based on thetayc distance metrics, comparing the initial to
the follow up samples, there was a significant difference between the
adenoma and carcinoma groups (P-value = 0.00198) {[}Figure 1a{]}. There
was also a significant difference in FIT between initial and follow up
samples with the carcinoma group having a significant decrease in FIT
versus the adenoma group (P-value = 2.15e-05) {[}Figure 1b{]}. The whole
community structure before and after surgery were visualized on NMDS
graphs for both adenoma {[}Figure1c{]} (PERMANOVA = 0.002) and carcinoma
{[}Figure 1d{]} (PERMANOVA = 0.997). When all initial and follow up
samples were compared to each other there was no significant overall
difference between them (PERMANOVA = 0.085). There was also no
significant difference between initial and follow up samples for
observed OTUs, Shannon diversity, or evenness after correction for
multiple comparisons {[}Table S1{]}. There was no significant difference
between initial and follow up samples for any single OTU {[}Figure
S1{]}.

@@

\textbf{\emph{Full and Reduced Model Results}} Since differences were
observed between initial and follow up samples and only a small number
of individuals were positive for previously associated CRC bacteria we
next investigated if we could create a model that could adequately
classify and adjust lesion probability based on the bacterial community
and FIT. The lesion model had an AUC range of 0.723 to 0.795 versus the
initial follow up model which had and AUC range of 0.451 to 0.67 after
100 20 repeated 10-fold cross validations. Interestingly, identification
of the most important variables and reducing the models to only these
factors increased the AUC in the lesion model (0.791 - 0.847) and
initial and follow up model (0.651 - 0.79).

@@

\textbf{\emph{Treatment and Time Differences}} There was no difference
in the amount of change in positive probability for either the full or
reduced lesion model for either chemotherapy (P-value = 0.821 and 0.821)
or radiation therapy (P-value = 0.69 and 0.981). Although the initial
follow up model was similar there was a significant decrease in positive
probability for those treated with chemotherapy (P-value = 7.04e-04 and
5.07e-03). Time elapsed between the collection of the follow up sample
from the initial sample, did not have a significant difference between
adenoma and carcinoma (uncorrected P-value = 0.784).

@@

There was no difference in alpha diversity metrics between the initial
and follow up samples {[}Table S1{]}. Yet based on thetayc there was a
significant difference between inital and follow up samples {[}Figure
1a{]} with carcinoma samples being more dissimilar to their initial
samples then adenoma samples. The change in FIT was also different
between initial and follow up for those with adenoma or carcinoma
{[}Figure 1b{]}. There was also a significant difference in the overall
bacterial community structure for those in the carcinoma group {[}Figure
1d{]} before and after surgery but not for the adenoma group {[}Figure
1c{]}. Investigation of the OTU relative abundance before and after
surgery found no single OTU to be significantly different {[}Figure
S1{]}.

Interestingly, when only previously associated cancer bacteria were
investigated only 2 out of the 4 had a significant decrease in relative
abundance between initial and follow up for the carcinoma group and 0
out of the 4 were significant for the adenoma group. This data suggests
that what may change after surgery is not necessarily any one specific
bacterium nor any addition of new or depressed bacterium but rather that
a shift of the bacterial community as a whole occurs with the existing
members of this microbiome.

We next created a model that incorporated FIT and the bacterial
microbiome to either be able to classify lesions (adenoma or carcinoma)
or initial and follow up samples in order to confirm that the community
was what was responding or chaning due to surgery. We found that the
OTUs that made up most important variables to the model overwhelmingly
belonged to comensal bacteria. With only the lesion model having a
single OTU from previously associated cancer bacteria (Porphyromonas
asaccharolytica). Using only these important OTUs and FIT both models
(lesion and initial and follow up) had a significantly decreased
positive probability of either lesion or being an initial sample on
follow up {[}Figure 4 \& S4{]}. Confirming the importance of the changes
of comensal bacteria to these classifications was that a total of 14
OTUs were common to both models and all belonged to regular residents of
our gut community.

There was no difference for the majority of models tested for
differences in positive probablity based on whether chemotherapy or
radiation was received. There was also no difference in the thetayc
distance metric based on length of time between initial and follow up
sample between adenoma and carcinoma. These results would indicate that
the findings described on specific to the surgical intervention and that
differences observed between carcinoma and adenoma samples can not be
simply attributed to collection time between samples.

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool {[}13,14{]}
by exploring what happens to the bacterial community after surgical
removal of a lesion. Based on previous work by Arthur, et al. {[}18{]}
it may not be surprising to see E.coli as one of the most important OTUs
and one that was common to both models. Interestingly, many of the most
important OTUs had taxonomic identification for resident gut microbes.
This could suggest that the bacterial community is one of the first
components that could change during the pathogenesis of disease. These
bacterial microbiome changes could be the first step in allowing more
inflammatory bacterium to gain a foothold within the colon {[}19{]}.

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. One potential explanation is that even
with surgery and a shift of the bacterial community these specific
bacteria still persisted within the colon. An alternative reason would
be that even if they were reduced in relative abundance they were still
present. Another potential explanation is that they were not present in
enough individuals to be able to classify those with and without disease
with a high degree of accuracy. Finally, it is possible that our Random
Forest models were able to get the same information from measures such
as FIT or other OTUs. It is also possible that all three of these
potential explanations could have played a role. Regardless, our
observations would suggest that an individual's resident bacteria have a
role to play in disease initiation and could change in a way that allows
predictive models to lower the positive probability of a lesion after
surgery. It should be stated that our study does not argue against the
importance of these CRC associated bacteria in the pathogenesis of
disease but rather that the models do not utilize these specific
bacteria for classification purposes (lesion or initial and follow up).
In fact, it is possible that these CRC associated bacteria are important
in the transition from adenoma to carcinoma and would be one explanation
as to why in our data we not only see high initial relative abundances,
in certain individuals, but also large decreases in relative abundance
in those with carcinoma but not in those with adenoma after surgery
{[}Figure 2{]}.

One limitation of our study is that we do not know whether individuals
who were still classified as positive by the lesion model eventually had
a subsequent CRC diagnosis. This information would help to strengthen
the case for our Random Forest based model keeping a number of
individuals above the cutoff threshold even though at follow up they
were diagnosed as no longer having a lesion. Another limitation is that
we do not know if adding modern tests such as the stool DNA test
{[}20{]} could help improve our overall AUC. Another limitation is that
this study drew heavily from those with caucasian ancestry. The results
may not be immediately representative of those with either Asian or
African ancestry. Finally, although our training and test set are
relatively large we still run the risk of overfitting or having a model
that may not be immediately extrapolateable to other populations. We've
done our best to safeguard against this by not only running 10-fold
cross validation but also having over 100 different 80/20 splits to try
and mimic the type of variation that might be expected to occur.

Within figure 3 the initial and follow up model showed better test AUC
results then the training set AUC. This may have occured because the
training AUC that was determined from 20 repeated 10 fold cross
validation removed samples at random and did not take into account that
they were matched samples. Another potential reason is that the model
itself may be overfit since the total number of samples is not that
large. However, the lesion model did not suffer from these discrepencies
and similar conclusions can be drawn solely from this model. Regardless,
further independent studies will need to be carried out to verify our
findings since not only are we dealing with stool, which could be very
different than the communities present on the actual tissue, but also
are dealing with correlations that may not be representative of the true
pathogensis of disease.

Despite these limitations we think that these findings significantly add
to the existing scientific knowledge on CRC and the bacterial
microbiome. The ability for machine learning algorithms to take
bacterial microbiome data and successfully lower positive probability
after either adenoma or carcinoma removal provides evidence that there
are specific signatures, mostly attributable to commensal organisms,
associated with these lesions. It also shows that these algorithms can
not only successfully react to successful treatment regimens but also
may be able to one day stratify CRC disease risk with a high level of
accuracy.

@@

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al
{[}13{]}. In brief, study exclusion involved those who had already
undergone surgery, radiation, or chemotherapy, had colorectal cancer
before a baseline stool sample could be obtained, had IBD, a known
hereditary non-polyposis colorectal cancer, or Familial adenomatous
polyposis. Samples used to build the model used for prediction were
collected either prior to a colonoscopy or between 1 - 2 weeks after.
The bacterial microbiome has been shown to nomralize within this time
period {[}21{]}. Kept apart from this training set were a total of 67
individuals that not only had a sample as described previoulsy but also
a follow up sample between 188 - 546 days after surgery and treatment
had been completed. This study was approved by the University of
Michigan Institutional Review Board. All study participants provided
informed consent and the study itself conformed to the guidelines set
out by the Helsinki Declaration.

\textbf{\emph{Fecal Immunochemical Test and 16S rRNA Gene Sequencing}}
FIT was analyzed as previously published using both OC FIT-CHEK and
OC-Auto Micro 80 automated system (Polymedco Inc.) {[}22{]}. 16S rRNA
gene sequencing was completed as previously described by Kozich, et al.
{[}23{]}. In brief, DNA extraction used the 96 well Soil DNA isolation
kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting
system (Eppendorf). The V4 variable region was amplified and the
resulting product was split between three sequencing runs with control,
adenoma, and carcinoma evenly represented on each run. Each group was
randomly assigned to avoid biases based on sample collection location.

\textbf{\emph{Sequence Processing}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described {[}23{]}. The general processing workflow
using mothur is as follows: Paired-end reads were first merged into
contigs, quality filtered, aligned to the SILVA database, screening for
chimeras, classified with a naive Bayesian classifier using the
Ribosomal Database Project (RDP), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm. The numer of sequences for each sample
was rarified to 10521 in an attempt to minimize uneven sampling.

\textbf{\emph{Lesion Model Creation}} The Random Forest {[}24{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) for the 67 individuals with follow up samples.
The model included data on FIT and the bacterial microbiome. Non-binary
data was checked for near zero variance and auto correlation. Data
columns that had near zero variance were removed. Columns that were
correlated with each other over a Spearman correlation coefficient of
0.75 had one of the two columns removed. This pre-processing was
performed with the R package caret (v6.0.73). Optimization of the mtry
hyperparameter involved taking the samples and making 100 80/20
(train/test) splits in the data where control and lesion were equally
represented in the 80 and 20 split, respectively. This 80\% portion was
then split again into an 80/20 split, and run through 20 repeated
10-fold cross validations to optimzie the model's AUC (Area Under the
Curve of the Receiver Operator Characteristic). This resulting model was
then tested on the 20\% of the data that was originally held out from
this overall process. Once the ideal mtry was found the entire 490
sample set was used to create the final Random Forest model on which
testing on the 67-person cohort was completed. The default cutoff of 0.5
was used as the threshold to classify individuals as positive or
negative for lesion. The hyperparameter, mtry, defines the number of
variables to investigate at each split before a new division of the data
is created.

\textbf{\emph{Initial Follow Up Model Creation}} We also investigated
whether a model could be created that could identify before and after
surgery samples. The training set utilized the 67-person cohort that was
previously used for testing of the lesion model. The creation of this
model and optimization of the mtry hyperparameter was completed using
the same procedure that was used to create the lesion model.

@@

\textbf{\emph{Statistical Analysis}} The R software package (v3.3.2) was
used for all statisitical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}25{]} in the vegan package
(v2.4.1) while comparisons between ROC curves utilized the method by
DeLong et al. {[}26{]} executed by the pROC (v1.8) package. Comparisons
between probabilities as well as overall amount of OTU between initial
and follow up samples utilized a paired wilcoxson ranked sum test. Where
multiple comparison testing was needed a Benjamini-Hochberg (BH)
correction was applied {[}27{]} and a corrected P-value of less than
0.05 was considered significant. Unless otherwise stated the P-values
reported are those of the BH corrected ones.

\textbf{\emph{Analysis Overview}} Differences in FIT between initial and
follow ups for either adenoma or carcinoma were investigated. Next,
initial and follow up samples were analyzed for differences in alpha and
beta diversity. All OTUs used in the lesion model were then also
analyzed using a paired wilcoxson test. We then investigated the
relative abundance of specific bacteria, that have been previously
associated with CRC, we selected OTUs that taxonomically classified to
Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus
Assacharolytica, and Porphyromonas Stomatis. Specifically, we wanted to
test if there were any differences based on whether the individual had
an adenoma or carcinoma. The lesion model was then tested for accuracy
in prediction and whether it reduced the positive probability of lesion
after surgery. The most important OTUs for this were used to build an
updated model and this reduced feature model was assessed for it's
similarity to the original model. We then used the initial follow up
model to assess whether this model could classify samples better then
the lesion model. The most important OTUs were then identified from this
model and used to create a reduced feature initial follow up model. This
reduced feature model, as was done with the lesion model, was compared
to the full model for loss of accuracy. Finally, a list of common OTUs
were found for the two different models used.

\textbf{\emph{Reproducible methods.}} A detailed and reproducible
description of how the data were processed and analyzed can be found at
\url{https://github.com/SchlossLab/Sze_followUps_2017}. Raw sequences
have been deposited into the NCBI Sequence Read Archive (SRP062005).

@@

\hypertarget{ref-arthur_intestinal_2012}{}
18. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
T-J, et al. Intestinal inflammation targets cancer-inducing activity of
the microbiota. Science (New York, N.Y.). 2012;338:120--3.

\hypertarget{ref-flynn_metabolic_2016}{}
19. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.

\hypertarget{ref-cotter_long-term_2016}{}
20. Cotter TG, Burger KN, Devens ME, Simonson JA, Lowrie KL, Heigh RI,
et al. Long-Term Follow-up of Patients Having False Positive
Multi-target Stool DNA Tests after Negative Screening Colonoscopy: The
LONG-HAUL Cohort Study. Cancer Epidemiology, Biomarkers \& Prevention: A
Publication of the American Association for Cancer Research, Cosponsored
by the American Society of Preventive Oncology. 2016;

\hypertarget{ref-obrien_impact_2013}{}
21. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.

\hypertarget{ref-daly_evaluation_2013}{}
22. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
for colorectal cancer screening. Journal of Primary Care \& Community
Health. 2013;4:245--50.

\hypertarget{ref-kozich_development_2013}{}
23. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq Illumina sequencing
platform. Applied and Environmental Microbiology. 2013;79:5112--20.

\hypertarget{ref-breiman_random_2001}{}
24. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
{[}cited 2013 Feb 7{]};45:5--32. Available from:
\href{http://link.springer.com/article/10.1023/A\%3A1010933404324\%20http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}{http://link.springer.com/article/10.1023/A\%3A1010933404324 http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}

\hypertarget{ref-anderson_permanova_2013}{}
25. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
the face of heterogeneous dispersions: What null hypothesis are you
testing? Ecological Monographs {[}Internet{]}. 2013 {[}cited 2017 Jan
5{]};83:557--74. Available from:
\url{http://doi.wiley.com/10.1890/12-2010.1}

\hypertarget{ref-delong_comparing_1988}{}
26. DeLong ER, DeLong DM, Clarke-Pearson DL. Comparing the areas under
two or more correlated receiver operating characteristic curves: A
nonparametric approach. Biometrics. 1988;44:837--45.
